---
title: "State Crops"
author: "Mandy Liesch"
date: "10/23/2021"
always_allow_html: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
``` 

```{css, echo=FALSE}
h1, h2, h3 {
  text-align: center;
}
```

```{r, echo=FALSE, message = FALSE, warning=FALSE}
library(tidyverse)  # data manipulation
library(rmarkdown)
library(usethis)
library(knitr)
library(RColorBrewer)
library(reshape2)
library(hrbrthemes)
library(plotly)
library(ggplot2)
use_git_config(user.name="Mandy Liesch", user.email="amliesch@ncsu.edu")
```

```{r, eval=FALSE, echo=FALSE}
#code to create the rendering. this chunk is put into the code to knit and
#render the function.
rmarkdown::render("StateAcresHeatmap.Rmd", 
                  output_format = "github_document", 
                  output_options = list(toc=TRUE, toc_depth = 3), 
                  output_file = "README.md")
```


## USDA NASS Data

```{r, echo = FALSE}
prePro<-function(crop){
  cropFile<-paste(crop,".csv", sep = "")
  return(cropFile)
}
```
State Crop Production Data from 1990-2021 in acres. The NASS data for each crop was downloaded by state, and trimmed into three columns in Microsoft Excel: Year, State, and Acres. The total land area of each state is noted in the GIS shapefiles. The square meter values were turned to acreages using the calculate function in an attribute table, and then the attribute table was exported to a .csv file for use in calculating the total crop area

```{r, echo = FALSE}
# Develop a function to read in cropping history from 
# 1990-2021 (or whatever year is latest)
cropRead<-function(crop){
  #####
  ## This function is designed to take the long form ###  
  ## using year and state, and compresses it to the  ###  
  ## wide form, with a state having an individual    ###  
  ## line, then it calculates the TOTAL percentage   ###    
  ## land area of each crop and returns a data frame ###
  ##### 
  cropFile<-prePro(crop)
  #read in the input CSV file.
  cropLong<-read.csv(cropFile)
  #create an acre output column.
  cropLong$acres<-cropLong[,3]
  #Default values tend to be characters with commas. This  
  # line ensures that the data is transformed to numeric.
  cropLong$acres<-as.numeric(gsub(",", "", cropLong$acres))
  #Remove the initial character column
  cropLong[,3]<-NULL
  #Transfer the crop acreage from long format to wide
  cropWide<-spread(cropLong, key = Year, value = c(3))
  # The state Acres file was derived from GIS polygons  
  # from the Census Bureau. Square meters of land area  
  # was converted into acres in ArcMap 10.4.1
  stateAcres<-read.csv("STATEAcres.csv")
  # Make sure the state name is changed to uppercase
  stateAcres$NAME<-toupper(stateAcres$NAME)
  # Merge the state acres data with the crop wide data  
  # by state name, only keeping those that appear in the  
  # cropping wide matrix. 
  cropMerge<-merge(stateAcres, cropWide, by.x = 'NAME', by.y = 'State', all.y=TRUE)
  # Calculate the percentage of the total crop area  
  # produced per year divided by the acreage.
  percentages<-  cropMerge %>%
   mutate_at(vars(4:35) , funs(P =./cropMerge$AcreArea * 100))
  
  # Remove the raw acreage values and leave only the
  # percentages
  cropNew <- percentages[, c(1, 36:67)]
  # re-add commodity name0
  cropNew$crop<-crop
  return(cropNew)
}
```

Once the file is read in and transformed, with new percentages calculated, the data is linked to a second function that transforms the data into a matrix format for use in the creation of heatmaps and plots. 

```{r, echo = FALSE}
heatProcess<-function(cropFile){
  cropNew<-cropRead(cropFile)
  # Use the states as row names
  # Remove the Name Variable

  cropNew$crop<-NULL
  
  cropNew<-filter(cropNew, NAME != "OTHER STATES")
  
  cropNew[is.na(cropNew)]<-0
  cropNew<- cropNew[order(cropNew$`1990_P`),]

  
  colnames(cropNew)<-c("NAME", "'90", "'91", "'92", "'93","'94", "'95","'96", "'97","'98", "'99","'00", "'01", "'02", "'03", "'04", "'05", "'06", "'07", "'08", "'09", "'10", "'11", "'12", "'13","'14", "'15","'16", "'17","'18", "'19", "'20", "'21")
  
  longFrame<-melt(cropNew, id.vars=c("NAME"))
  
  longFrame <- longFrame %>%
  mutate(text = paste0("State: ", NAME, "\n", "Year: ", variable, "\n", "Percentage: ",round(value,2), "\n"))
  
  p<-  ggplot(longFrame, aes(variable, NAME, fill= value, text=text)) + 
    geom_tile() +
    scale_fill_gradient(low="white", high="red") +
    theme_ipsum() +
    theme(legend.position="right")
  
  finalPlot<-ggplotly(p, tooltip="text")
  
  return(finalPlot)
}

```

And this function uses the data matrix from the above functions, and transfers the matrix dataframe to a graphical heat map based on the crop selected.


```{r, echo=FALSE}
#create a megafile with all state and crop information

allCrops<-c('alfalfaHay', 'barley', 'canola', 'corn', 'cotton', 'drybeans', 'durumwheat', 'flaxseeds', 'hay', 'hops', 'lentils', 'oats', 'peanuts', 'peas', 'potatos', 'rice', 'sorghum', 'soybean', 'springwheat', 'sugarbeets', 'sugarcane', 'sunflower', 'sweetpotato', 'tobacco', 'winterwheat')

finalCropFrame<-lapply(X = allCrops, FUN = cropRead)

cropFrame<-data.frame(Reduce(rbind, finalCropFrame))

cropFrame[is.na(cropFrame)]<-0

```

```{r,  echo=FALSE}
stateCrop<-function(state){
  stateNew<-cropFrame
  stateName<-state
  filterState<-filter(stateNew,NAME==stateName)
  filterState$NAME<-NULL
  
  colnames(filterState)<-c("'90", "'91", "'92", "'93","'94", "'95","'96", "'97","'98", "'99","'00", "'01", "'02", "'03", "'04", "'05", "'06", "'07", "'08", "'09", "'10", "'11", "'12", "'13","'14", "'15","'16", "'17","'18", "'19", "'20", "'21", "crop")
  
  longFrame<-melt(filterState, id.vars=c("crop"))
  
  longFrame<- longFrame[order(longFrame$crop),]
  
  longFrame <- longFrame %>%
  mutate(text = paste0("Crop: ", crop, "\n", "Year: ", variable, "\n", "Percentage: ",round(value,2), "\n"))

  p<-  ggplot(longFrame, aes(variable, crop, fill= value, text=text)) + 
    geom_tile() +
    scale_fill_gradient(low="white", high="red") +
    theme_ipsum() +
    theme(legend.position="right")
  
  finalPlot<-ggplotly(p, tooltip="text")
  
  return(finalPlot)
}



```

## Key State Information

### North Dakota
```{r,  echo=FALSE}
ND<-stateCrop("NORTH DAKOTA")
ND
```

### Wisconsin
```{r,  echo=FALSE}
WI<-stateCrop("WISCONSIN")
WI
```

### Iowa
```{r,  echo=FALSE}
IA<-stateCrop("IOWA")
IA
```


## Major Cash Crops

### Corn
```{r, echo = FALSE, fig.align ='center'}
corn<-heatProcess('corn')
corn
```

### Soybean
```{r, echo = FALSE, fig.align ='center'}
soybean<-heatProcess('soybean')
soybean
```

### Cotton
```{r, echo = FALSE, fig.align ='center'}
cotton<-heatProcess('cotton')
cotton
```

### Rice
```{r, echo = FALSE, fig.align ='center'}
rice<-heatProcess('rice')
rice
```

## Wheat

### Winter Wheat
```{r, echo = FALSE, fig.align ='center'}
wintWheat<-heatProcess('winterwheat')
wintWheat
```

### Spring Wheat
```{r, echo = FALSE, fig.align ='center'}
sprWheat<-heatProcess('springwheat')
sprWheat
```

### Durum Wheat
```{r, echo = FALSE, fig.align ='center'}
durwheat<-heatProcess('durumwheat')
durwheat
```

## Other Grains
### Oats
```{r, echo = FALSE, fig.align ='center'}
oats<-heatProcess('oats')
oats
```

### Barley
```{r, echo = FALSE, fig.align ='center'}
barley<-heatProcess('barley')
barley
```

### Flaxseed
```{r, echo = FALSE, fig.align ='center'}
flaxseed<-heatProcess('flaxseeds')
flaxseed
```

### Sorghum
```{r, echo = FALSE, fig.align ='center'}
sorghum<-heatProcess('sorghum')
sorghum
```

## Oil Crops

### All Sunflower
```{r, echo = FALSE, fig.align ='center'}
sunflower<-heatProcess('sunflower')
sunflower
```

### Canola
```{r, echo = FALSE, fig.align ='center'}
canola<-heatProcess('canola')
canola
```

## Sugar Crops
### Sugar Beets
```{r, echo = FALSE, fig.align ='center'}
sugarbeets<-heatProcess('sugarbeets')
sugarbeets
```

### Sugarcane (Harvested Sugar and Seed)
```{r, echo = FALSE, fig.align ='center'}
sugarcane<-heatProcess('sugarcane')
sugarcane
```

## Potatoes
### Potatoes
```{r, echo = FALSE, fig.align ='center'}
potato<-heatProcess('potatos')
potato
```

### Sweet Potatoes (No 2021 Data)
```{r, echo = FALSE, fig.align ='center'}
sweetPotato<-heatProcess('sweetpotato')
sweetPotato
```

## Pulses and Beans
### Dry Beans
```{r, echo = FALSE, fig.align ='center'}
drybeans<-heatProcess('drybeans')
drybeans
```

### Dry Peas (And Austrian Peas)
```{r, echo = FALSE, fig.align ='center'}
peas<-heatProcess('peas')
peas
```

### Lentils
```{r, echo = FALSE, fig.align ='center'}
lentils<-heatProcess('lentils')
lentils
```

### Peanuts
```{r, echo = FALSE, fig.align ='center'}
peanuts<-heatProcess('peanuts')
peanuts
```

## Other Crops
### Tobacco (Harvested Acres ONLY)
```{r, echo = FALSE, fig.align ='center'}
tobacco<-heatProcess('tobacco')
tobacco
```

### Hops
```{r, echo = FALSE, fig.align ='center'}
hops<-heatProcess('hops')
hops
```

## Hay

### All Hay
```{r, echo = FALSE, fig.align ='center'}
hay<-heatProcess('hay')
hay
```

### Alfalfa Hay
```{r, echo = FALSE, fig.align ='center'}
alfalfa<-heatProcess('alfalfaHay')
alfalfa
```
